WASP-4 is Accelerating Towards the Earth

### 2020-02-03 paper/ms.tex
tweaked tables

\begin{abstract}
  Space and ground-based transit measurements of the hot Jupiter
  WASP-4b have recently shown that its orbital period appears to be
  decreasing.  Proposed explanations for the period change include
  tidal orbital decay, orbital precession, and light-travel time
  effects.  We present new radial velocity measurements of WASP-4,
  acquired with Keck-HIRES.  The data show that the system is
  accelerating towards the Earth at $\dot{\gamma} =
  -0.0422^{+0.0028}_{-0.0027}\,{\rm m}\,{\rm s}^{-1}\,{\rm day}^{-1}$.
  The implied period decrease explains most or all of WASP-4's
  changing orbital period as a light-travel time effect ($-5.94 \pm
  0.39~{\rm msec}\,{\rm yr}^{-1}$ implied, compared to $-8.64 \pm
  1.26~{\rm msec}\,{\rm yr}^{-1}$ observed).  Combining the radial
  velocities with new upper limits from speckle imaging, we find that
  the system's acceleration is likely caused by a 10-200$\,M_{\rm
  Jup}$ companion with semi-major axis between 10-100$\,$AU.  The
  statistics from \citet{knutson_friends_2014} imply that about 1 in 6
  hot Jupiters are expected to show comparable period decreases to
  WASP-4, due to acceleration by outer companions.  These
  period decreases will be increasingly measureable in coming years,
  because the precision of the period derivative measurement scales with
  the observing baseline to the power of 5/2.  Continued radial velocity monitoring
  of hot Jupiters is therefore essential to distinguish tidal orbital
  decay from line-of-sight accelerations.
\end{abstract}

@@

We fitted each of the two models by assuming a Gaussian likelihood and
sampling over the posterior probability distributions.  We sampled the
posterior using the algorithm proposed by
\citet{goodman_ensemble_2010} and implemented by
\citet{foreman-mackey_emcee_2013} in \texttt{emcee}.  The prior for
the quadratic model allowed the period derivative to have any sign.

@@

The resulting best-fit period derivative for the quadratic model is 
\begin{equation}
\dot{P}
  = - (2.74 \pm 0.28)\times 10^{-10}
  = - 8w.64 \pm 1.26~{\rm msec}\,{\rm yr}^{-1}.
  \label{eq:dP_dt_obs}
\end{equation}
This agrees to within $1\sigma$ of the value reported by
\citet{southworth_transit_2019} ($\dot{P} = - 9.2 \pm 1.1~{\rm
ms}\,{\rm yr}^{-1}$).  It is $\approx$2.3$\sigma$ larger than the rate
of period decrease we previously reported \citepalias[$- 12.6 \pm 1.2~{\rm
ms}\,{\rm yr}^{-1}$;][]{bouma_wasp4b_2019}, presumably because of the
new data from \citeauthor{southworth_transit_2019} and
\citeauthor{baluev_2019}  The other best-fit transit timing model
parameters are reported in Table~3.

\subsection{Radial Velocities: WASP-4's acceleration towards the Earth}

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.7\textwidth]{f2.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{
    {\bf Radial velocities of WASP-4.}
    {\it Top:} RV measurements, with best-fit instrument offsets
    added. 
    {\it Middle:}
    Residuals, after subtracting the
    maximum-likelihood orbit of WASP-4b.  The linear trend
    inferred from the RV data is shown with a black line.
    The trend that would be needed to
    produce the period decrease seen in transits ($\dot{P}= - 8.64 \pm
    1.26~{\rm msec}\,{\rm yr}^{-1}$) is indicated in purple.  The four
    new RV measurements from this work increase the
    significance of the linear trend from $\approx$2$\sigma$
    to 15$\sigma$.
    {\it Bottom:}
    Phased orbit of WASP-4b.
	\label{fig:rvs}
  \vspace{-0.3cm}
	}
\end{figure*}

@@

In the preferred ``planet+linear trend'' model, WASP-4 is accelerating
towards our line-of-sight at high confidence,
\begin{equation}
  \dot{v}_{\rm r} = \dot{\gamma} = 
     -0.0422^{+0.0028}_{-0.0027}
     \,{\rm m}\,{\rm s}^{-1}\,{\rm day}^{-1}.
\end{equation}
For comparison, before our new measurements, $\dot{\gamma}$ was
thought to be about five times smaller, and had marginal statistical
significance \citep{knutson_friends_2014,bouma_wasp4b_2019}.

@@

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.8\textwidth]{f4.pdf}
	\end{center}
	\vspace{-0.6cm}
	\caption{
    {\bf Masses and semi-major axes of companions that meet
    requirements of both the radial velocities and the speckle
    imaging.} The likelihood inferred from radial velocities is shown
    in grayscale, and the region excluded from the speckle imaging is
    indicated.  The expected mass to semimajor-axis degeneracy is
    shown with a black line.
	\label{fig:mass_sma}
  \vspace{-0cm}
	}
\end{figure*}

@@

To evaluate the importance of these effects for future transit timing
analyses, we collected the linear radial velocity trends reported by
\citet{knutson_friends_2014}, and computed the expected orbital period
derivatives $\dot{P}_{\rm\,RV} = \dot{v}_{\rm r} P/c$ for each system.
The results are given in Table~4, and visualized for hot Jupiters with
significant ($>$$3\sigma$) linear trends in Figure~\ref{fig:pdot_pop}.

@@

\input{WASP-4b_transit_time_table.tex}
\vspace{-2cm}
\input{WASP-4b_rv_table.tex}
\vspace{-2cm}
\input{model_fit_table.tex}
\vspace{-2cm}
\input{pdot_table.tex}

>>>>

\begin{abstract}
  Space and ground-based transit measurements of the hot Jupiter
  WASP-4b have recently shown that its orbital period appears to be
  decreasing.  Proposed explanations for the period change include
  tidal orbital decay, orbital precession, and light-travel time
  effects.  We present new radial velocity measurements of WASP-4,
  acquired with Keck-HIRES.  The data show that the system is
  accelerating towards the Earth at $\dot{\gamma} =
  -0.0422^{+0.0028}_{-0.0027}\,{\rm m}\,{\rm s}^{-1}\,{\rm day}^{-1}$.
  The implied period decrease explains most or all of WASP-4's
  changing orbital period as a light-travel time effect ($-5.94 \pm
  0.39~{\rm msec}\,{\rm yr}^{-1}$ implied, compared to $-8.64 \pm
  1.26~{\rm msec}\,{\rm yr}^{-1}$ observed).  Combining the radial
  velocities with new upper limits from speckle imaging, we find that
  the system's acceleration is likely caused by a 10-200$\,M_{\rm
  Jup}$ companion with semi-major axis between 10-100$\,$AU.  The
  statistics from \citet{knutson_friends_2014} imply that about 1 in 6
  hot Jupiters are expected to show period decreases comparable to
  WASP-4, due to acceleration by outer companions.  These period
  changes will be increasingly measureable in coming years, because
  the precision of the period derivative measurement scales with the
  observing baseline to the power of 5/2.  Continued radial velocity
  monitoring of hot Jupiters is therefore essential to distinguish
  tidal orbital decay from line-of-sight accelerations.
\end{abstract}

@@

We fitted each model by assuming a Gaussian likelihood and
sampling over the posterior probability distributions.  We sampled the
posterior using the algorithm proposed by
\citet{goodman_ensemble_2010} and implemented by
\citet{foreman-mackey_emcee_2013} in \texttt{emcee}.  The prior for
the quadratic model allowed the period derivative to have any sign.

@@

The resulting best-fit period derivative for the quadratic model is 
\begin{equation}
\dot{P}
  = - (2.74 \pm 0.28)\times 10^{-10}
  = - 8.64 \pm 1.26~{\rm msec}\,{\rm yr}^{-1}.
  \label{eq:dP_dt_obs}
\end{equation}
This agrees to within $1\sigma$ of the value reported by
\citet{southworth_transit_2019} ($\dot{P} = - 9.2 \pm 1.1~{\rm
ms}\,{\rm yr}^{-1}$).  It is $\approx$2.3$\sigma$ larger than the rate
of period decrease we previously reported \citepalias[$- 12.6 \pm 1.2~{\rm
ms}\,{\rm yr}^{-1}$;][]{bouma_wasp4b_2019}, presumably because of the
new data from \citeauthor{southworth_transit_2019} and
\citeauthor{baluev_2019}  The other best-fit transit timing model
parameters are reported in Table~3.

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.7\textwidth]{f2.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{
    {\bf Radial velocities of WASP-4.}
    {\it Top:} RV measurements, with best-fit instrument offsets
    added. 
    {\it Middle:}
    Residuals, after subtracting the
    maximum-likelihood orbit of WASP-4b.  The linear trend
    inferred from the RV data is shown with a black line.
    The trend that would be needed to
    produce the period decrease seen in transits ($\dot{P}= - 8.64 \pm
    1.26~{\rm msec}\,{\rm yr}^{-1}$) is indicated in purple.  The four
    new RV measurements from this work increase the
    significance of the linear trend from $\approx$2$\sigma$
    to 15$\sigma$.
    {\it Bottom:}
    Phased orbit of WASP-4b.
	\label{fig:rvs}
  \vspace{-0.0cm}
	}
\end{figure*}

@@

In the preferred ``planet+linear trend'' model, WASP-4 is accelerating
towards our line-of-sight at high confidence,
\begin{equation}
  \dot{v}_{\rm r} = \dot{\gamma} = 
     -0.0422^{+0.0028}_{-0.0027}
     \,{\rm m}\,{\rm s}^{-1}\,{\rm day}^{-1}.
\end{equation}
The remaining radial velocity model parameters are listed in Table~4.
For comparison, before our new measurements, $\dot{\gamma}$ was
thought to be about five times smaller, and had marginal statistical
significance \citep{knutson_friends_2014,bouma_wasp4b_2019}.

@@

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.8\textwidth]{f4.pdf}
	\end{center}
	\vspace{-0.8cm}
	\caption{
    {\bf Masses and semi-major axes of companions that meet
    requirements of both the radial velocities and the speckle
    imaging.} The likelihood inferred from radial velocities is shown
    in grayscale, and the region excluded from the speckle imaging is
    indicated.  The expected mass to semimajor-axis degeneracy is
    shown with a black line.
	\label{fig:mass_sma}
  \vspace{-0cm}
	}
\end{figure*}

@@

To evaluate the importance of these effects for future transit timing
analyses, we collected the linear radial velocity trends reported by
\citet{knutson_friends_2014}, and computed the expected orbital period
derivatives $\dot{P}_{\rm\,RV} = \dot{v}_{\rm r} P/c$ for each system.
The results are given in Table~5, and visualized for hot Jupiters with
significant ($>$$3\sigma$) linear trends in Figure~\ref{fig:pdot_pop}.

@@

\input{WASP-4b_transit_time_table.tex}
\input{WASP-4b_rv_table.tex}
\input{model_fit_table.tex}
\input{rv_model_posterior_table.tex}
\input{pdot_table.tex}

### 2020-02-03 paper/pdot_table.tex
add publication-ready Pdot and RV tables
tweaked tables

\startlongtable
\begin{deluxetable*}{lllllllll}
%
\tablecaption{
	Predicted hot Jupiter period changes from linear radial
	velocity trends reported by \citet{knutson_friends_2014}.
\label{tab:pdot_table}
}
%
\tablenum{4}
%
% planet;gammadot;gammadot_pluserr;gammadot_minuserr;comment;pl_name;pl_orbper;Pdot;Pdot_upper_limit;Pdot_lower_limit;Pdot_perr;Pdot_merr;abs_Pdot;K14_significant
\tablehead{
  \colhead{Planet} &
  \colhead{$\dot{\gamma}$ [m$\,$s$^{-1}$$\,$yr$^{-1}$]} &
  \colhead{$+\sigma_{\dot{\gamma}}$ [m$\,$s$^{-1}$yr$^{-1}$]} & 
  \colhead{$-\sigma_{\dot{\gamma}}$ [m$\,$s$^{-1}$yr$^{-1}$]} & 
  \colhead{$P$ [days]} &
  \colhead{$\dot{P}_{\,{\rm RV}}$ [ms$\,$yr$^{-1}$]} &
  \colhead{$+\sigma_{\dot{P}_{\,{\rm RV}}}$ [ms$\,$yr$^{-1}$]} &
  \colhead{$-\sigma_{\dot{P}_{\,{\rm RV}}}$ [ms$\,$yr$^{-1}$]} &
  \colhead{Significant?}
}
%
%FIXME precision
\startdata
HAT-P-2 b & -0.0938 & 0.0067 & 0.0069 & 5.6335158 & -55.62 & 3.97 & 4.09 & 1
\enddata
%
\tablecomments{
  Table~4 is published in its entirety in a machine-readable format.
  A single row is shown here for guidance regarding form and content.
}
%
\end{deluxetable*}

>>>>

\startlongtable
\begin{deluxetable*}{lllllllll}
%
\tablecaption{
	Predicted hot Jupiter period changes from linear radial
	velocity trends reported by \citet{knutson_friends_2014}.
\label{tab:pdot_table}
}
%
\tablenum{5}
%
% planet;gammadot;gammadot_pluserr;gammadot_minuserr;comment;pl_name;pl_orbper;Pdot;Pdot_upper_limit;Pdot_lower_limit;Pdot_perr;Pdot_merr;abs_Pdot;K14_significant
\tablehead{
  \colhead{Planet} &
  \colhead{$\dot{\gamma}$ [m$\,$s$^{-1}$$\,$yr$^{-1}$]} &
  \colhead{$+\sigma_{\dot{\gamma}}$ [m$\,$s$^{-1}$yr$^{-1}$]} & 
  \colhead{$-\sigma_{\dot{\gamma}}$ [m$\,$s$^{-1}$yr$^{-1}$]} & 
  \colhead{$P$ [days]} &
  \colhead{$\dot{P}_{\,{\rm RV}}$ [ms$\,$yr$^{-1}$]} &
  \colhead{$+\sigma_{\dot{P}_{\,{\rm RV}}}$ [ms$\,$yr$^{-1}$]} &
  \colhead{$-\sigma_{\dot{P}_{\,{\rm RV}}}$ [ms$\,$yr$^{-1}$]} &
  \colhead{Significant?}
}
%
%FIXME precision
\startdata
HAT-P-2 b & -0.0938 & 0.0067 & 0.0069 & 5.6335158 & -55.62 & 3.97 & 4.09 & 1
\enddata
%
\tablecomments{
  Table~5 is published in its entirety in a machine-readable format.
  The first entry is shown for guidance regarding form and content.
  Orbital periods were retrieved from NASA's Exoplanet Archive.
  Additional comments regarding non-linear trends and stellar activity
  are included in the MRT.
}
%
\end{deluxetable*}
